Bayesian Rate Estimation

### 2012-08-14 many.tex
Started to add in theory relating our work to loudest event etc.

@@

\author{Jonathan R. Gair}
\email{jrg23@cam.ac.uk}

@@

Putting everything together, the posterior is
\begin{equation}
  \label{eq:posterior}
  p\left( \mathset{f_i}, R_f, R_b, \theta | d \right) \propto \left[
    \prod_{\mathset{i | f_i = 1}} R_f \hat{f}\left(x_i, \theta\right) \right]
  \left[ \prod_{\mathset{i | f_i = 0}} R_b \hat{b}\left( x_i, \theta\right)
  \right] \exp\left[ - \left(R_f + R_b\right) \right]
  \frac{p(\theta)}{\sqrt{R_f R_b}}
\end{equation}
Not surprisingly, this posterior is equivalent to one derived by
assuming that the flags, $\mathset{f_i}$, are un-observed data and
treating the sets $\mathset{x_i | f_i = 1}$ and $\mathset{x_i | f_i =
  0}$ as samples from an inhomogeneous Poisson process.  For an
inhomogeneous Poisson process with rate function $r(y)$ (cumulative
rate $R(y)$), the likelihood of a set of samples $\mathset{y_i}$ is
given by 
\begin{multline}
  p\left( \mathset{y_i} | r \right) = \lim_{\epsilon \to 0} P\left(
  \textnormal{zero events below } y_0 - \epsilon \right) \times
  P\left( \textnormal{one event between } y_0 - \epsilon \textnormal{
    and } y_0 + \epsilon \right) \\ \times P\left( \textnormal{zero
    events between } y_0+\epsilon \textnormal{ and } y_1 - \epsilon
  \right) \ldots \\ = \lim_{\epsilon \to 0} \exp\left[ - R\left(y_0 -
    \epsilon \right) \right] \left[ r\left( y_0 \right) +
    \order{\epsilon^2} \right] \times \exp\left[ - \left[ R\left( y_1
      - \epsilon\right) - R\left( y_0 + \epsilon\right) \right]
    \right] \times \ldots = \left[\prod_i r\left( y_i
    \right)\right] \exp\left[ - R\left( \infty \right) \right].
\end{multline}
\ilya{[Note: I don't think this is right -- the expression you've written down should be zero in the limit $\epsilon \to 0$, and it would be if the $\epsilon$s were kept correctly (e.g., $ P\left( \textnormal{one event between } y_0 - \epsilon \textnormal{ and } y_0 + \epsilon \right) \sim  r\left( y_0 \right) \epsilon +
    \order{\epsilon^2}$).]}
Applying this likelihood once to the foreground samples and once to
the background samples reproduces Eq.~\eqref{eq:posterior}.

We can marginalize the posterior over the flags, $f_i$, obtaining
\begin{equation}
  \label{eq:rate-shape-posterior}
  p\left( R_f, R_b, \theta | d \right) = \sum_{\mathset{f_i} \in
    \mathset{0,1}^N} p\left( \mathset{f_i}, R_f, R_b, \theta | d \right)
  \propto \prod_{i} \left[ R_f \hat{f}\left(x_i, \theta\right) + R_b
    \hat{b}\left( x_i, \theta\right) \right] \exp\left[-\left( R_f +
      R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
\ilya{[Note: some of the expressions here aren't really properly normalized -- should check carefully which ones are, and replace $=$ by $\propto$ for the others.]}
This expression is useful if we are only interested in rates and not
the probability that any particular event is foreground or background.
Unlike the full posterior (Eq.~\eqref{eq:posterior}),
Eq.~\eqref{eq:rate-shape-posterior} contains only continuous
parameters.

@@

>>>>

\author{Jonathan R. Gair}
\email{jrg23@cam.ac.uk}
\affiliation{Institute of Astronomy\\University of Cambridge\\Madingley Road, Cambridge CB3 0HA\\United Kingdom}

@@

@@

Putting everything together, the posterior is
\begin{equation}
  \label{eq:posterior}
  p\left( \mathset{f_i}, R_f, R_b, \theta | d \right) \propto \left[
    \prod_{\mathset{i | f_i = 1}} R_f \hat{f}\left(x_i, \theta\right) \right]
  \left[ \prod_{\mathset{i | f_i = 0}} R_b \hat{b}\left( x_i, \theta\right)
  \right] \exp\left[ - \left(R_f + R_b\right) \right]
  \frac{p(\theta)}{\sqrt{R_f R_b}}
\end{equation}
Not surprisingly, this posterior is equivalent to one derived by
assuming that the flags, $\mathset{f_i}$, are un-observed data and
treating the sets $\mathset{x_i | f_i = 1}$ and $\mathset{x_i | f_i =
  0}$ as samples from an inhomogeneous Poisson process.  For an
inhomogeneous Poisson process with rate function $r(y)$ (cumulative
rate $R(y)$), the likelihood of a set of samples $\mathset{y_i}$ is
given by 
\begin{multline}
  p\left( \mathset{y_i} | r \right) \,{\rm d}^Ny_i= P\left(
  \textnormal{zero events below } y_0  \right) \times
  P\left( \textnormal{one event between } y_0  \textnormal{
    and } y_0 + {\rm d}y_0 \right) \\ \times P\left( \textnormal{zero
    events between } y_0+{\rm d}y_0 \textnormal{ and } y_1
  \right) \ldots \\ p\left( \mathset{y_i} | r \right) = \lim_{\delta y_i \to 0} \exp\left[ - R\left(y_0\right) \right] \left[ r\left( y_0 \right) +
    \order{\delta y_0} \right] \times \exp\left[ - \left[ R\left( y_1
      \right) - R\left( y_0 + \delta y_0\right) \right]
    \right] \times \ldots \\ = \left[\prod_i r\left( y_i
    \right)\right] \exp\left[ - R\left( \infty \right) \right].
\end{multline}
Applying this likelihood once to the foreground samples and once to
the background samples reproduces Eq.~\eqref{eq:posterior}.

We can marginalize the posterior over the flags, $f_i$, obtaining
\begin{equation}
  \label{eq:rate-shape-posterior}
  p\left( R_f, R_b, \theta | d \right) = \sum_{\mathset{f_i} \in
    \mathset{0,1}^N} p\left( \mathset{f_i}, R_f, R_b, \theta | d \right)
  \propto \prod_{i} \left[ R_f \hat{f}\left(x_i, \theta\right) + R_b
    \hat{b}\left( x_i, \theta\right) \right] \exp\left[-\left( R_f +
      R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
\ilya{[Note: some of the expressions here aren't really properly normalized -- should check carefully which ones are, and replace $=$ by $\propto$ for the others.]}
This expression is useful if we are only interested in rates and not
the probability that any particular event is foreground or background.
Unlike the full posterior (Eq.~\eqref{eq:posterior}),
Eq.~\eqref{eq:rate-shape-posterior} contains only continuous
parameters. We note that the terms that depend on the overall rate parameters, $R_b$ or $R_f$, are of the form $R_b^{n-1/2} \exp(-R_b)$ and so marginalisation over either $R_b$ or $R_f$ can be achieved analytically using
\begin{equation}
I_n = \int_0^{\infty} x^{n-\frac{1}{2}} \, {\rm e}^{-x} {\rm d} x = \frac{(2n-1)!!}{2^n} \sqrt{\pi}
\end{equation}
using the usual notation $(2n-1)!! \equiv (2n-1)(2n-3)\cdots1$.

@@

It is informative to relate these results to two other ways to estimate the foreground rate parameter --- the loudest event statistic and the foreground-dominated statistic.

\subsubsection{Loudest event statistic}
If we were to include only the $k$-loudest events in the posterior distribution, rather than all observed events, the posterior (Eq.~\eqref{eq:posterior}) would be modified by an additional factor of $\exp[-(R_f \hat{F}(x_{N-k+1}) + R_b \hat{B}(x_{N-k+1}))]$, where we have assumed events are ordered by loudness, so that $x_{N-k+1}$ is the $k$-th loudest event. This term accounts for the data-dependent threshold that a loudest event statistic employs.

For the usual $k=1$ case, the marginalised posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{equation}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f \hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right)\exp\left[-\left( R_f (1-\hat{F}(x_N,\theta)) +
      R_b (1-\hat{B}(x_N,\theta)) \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
where $x_N$ denotes the loudness of the loudest event. Marginalising over $R_b$ we obtain
\begin{equation}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left( \frac{\hat{b}(x_N,\theta)}{2(1-\hat{B}(x_N,\theta))} + R_f \hat{f}(x_N,\theta)\right)\frac{\sqrt{\pi}}{\sqrt{1-\hat{B}(x_N,\theta)}\,\sqrt{R_f}} \exp(-R_f) .
\end{equation}
This posterior has a maximum in $R_f$ at
\begin{equation}
R_f = \frac{\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta)\tilde{b}(x_N,\theta) \pm \sqrt{(\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta)\tilde{b}(x_N,\theta))^2 - 4\tilde{b}(x_N,\theta)(1-\hat{F}(x_N,\theta))\hat{f}(x_N,\theta)}}{4 \hat{f}(x_N,\theta)(1-\hat{F}(x_N,\theta)}.
\end{equation}
where $\tilde{b}(x_N,\theta) =\hat{b}(x_N,\theta)/(1-\hat{B}(x_N,\theta))$. If we assume that $\tilde{b}(x_N,\theta) \ll 1$, i.e., that the loudest event is very likely to be foreground we obtain the expected result $(1- \hat{F}(x_N,\theta))\,R_f \approx 1/2$.
